---
title: "SVAlignR and Virus-Associated Cancer"
author: "Kevin R. Coombes, Keiko Akagi, Davied Symer, and Maura Gillison"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    theme: journal
    highlight: kate
  toc: true
vignette: >
  %\VignetteIndexEntry{SVAlignR and Virus-Associated Cancer}
  %\VignetteKeywords{OOMPA,CytoGPS,karyotypes,idiograms,graphics}
  %\VignetteDepends{SVAlignR}
  %\VignettePackage{SVAlignR}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r opts, echo=FALSE, results="hide"}
knitr::opts_chunk$set(fig.width=8, fig.height=5)
options(width = 88)
```
```{r mycss, results="asis", echo=FALSE}
cat('
<style type="text/css">
.figure { text-align: center; }
.caption { font-weight: bold; }
</style>
')
```
# Introduction
Some cancers are associated with the presence of viruses that disrupt chromosomal DNA.
Among the best known examples are cervical cancers and head-and-neck cancers that are
associated with human papilloma virus (HPV). Using whole genome short-read sequencing,
it is possible to identify breakpoints, including those where viral DNA is merged with
human DNA. It remains an open question whether those fragments are fully integrated into
a human chromosome or just exist as circular extrachromosomal DNA. Using newer long-read
sequencing technology, one can obtain information about the order of multiple breakpoints
in the original DNA molecules in the tumor sample. In this vignette, we describe SVAlignR,
an R package intended to use these long-read sequences to reconstruct those original molecules.

# Getting Started
We begin by loading the `SVAlignR` package, and a sample data set.
```{r Setup}
library(SVAlignR)
data(longreads)
head(longreads[, 1:3])
```
Each of the  `r nrow(longreads)` rows in this spreadsheet represents a different Oxford
Nanopore long read.  The `qname` column is a unique identifier; the `qlen` column contains
the length of the read in bytes; and the `connections` column lists the breakpoints that 
he read contains, each of which has been assigned a numeric identifier. We are interested
in this column.  We have two main goals:

1. Assemble the long-reads (which should be thought of as fragmentary) into complete molecules.
2. Start constructing a phylogenetic tree showing the evolution of these HPV-human combining
molecules.

As with many sequence alignment problems, we start by removing any duplicate "reads"; here
that means duplication of the break-point sequence summaries of the actual reads.
```{r LR}
LR0 <- longreads$connection
names(LR0) <- rownames(longreads)
LR <- LR0[!duplicated(LR0)]
weights <- sapply(LR, function(lr) sum(LR0 == lr))
length(LR) # expect 158
```

# Clustering Sequences
We want to cluster the break-point sequences.

## Recoding
In order to make use of existing algorithms,
we need to convert the numeric identifiers into single characters as part of an "alphabet". 
The `Cipher` class takes in a character vector containing the "words" (that is, our
break-point sequences) that need to be rewritten. By default, it assumes (as in our example)
that "letters" are separated by hyphens, identifies the unique symbols, and creates a tool
to convert back-and-forth from one "alphabet" to another.
```{r unique.symbols}
alphabet <- Cipher(LR)
alphabet
```

Now we are able to convert the original sequences of numbers into words in our specialized
alphabet.
```{r pseudo}
head(LR)
pseudo <- encode(alphabet, LR)
head(pseudo)
```

## Clusters
The recoding step is includes inth e `SequenceCluster` constructor and will happen automatically.
So, to cluster the long-read break point sequences, we just need to cal the constructor on the
(named and de-duplicated) sequences. The core algorithm is the classic
Needelman-Wunsch global alignment algorithm, as implemented in the `NameNeedle` R package. As
used in the `SVAlignR` package, it computes an alignment score between any two sequences based
on a match score of 2, a mismatch penalty of -6, and a gap penalty of -2.  We fill a square
matrix with scores for every pair (I, J) of sequences. The resulting scores can be either
positive or negative. For each row X, we rescale into the interval [0,1] by computing
$$Y_{IJ} = 1 - \frac{X_{IJ} - min(X_{I.})}{max(X_{I.})-min(X_{I.})}.$$
Since the scores are not symmetric, we force symmetry by averaging the scaled score matrix and
its transpose.

We then perform hierarchical clustering using the symmetrized, rescaled score matrix to define
distances. We cut the dendrogam to cluster the sequences into 20 groups. The number 20 is
arbitrary. In our data set, it should yield about 8 long-read break-point sequences per cluster.
For later steps, we also want to make sure that each cluster contains at leeast 2 sequences.
```{r compute.distances}
seqclust <- SequenceCluster(LR, NC = 20)
table(seqclust@clusters)
```
Now we see that 20 clusters is appropriate. Each cluster is a relatively small size, but they
all contain more than one sequence so that aligning those sequences will be is meaningful.
**Figure 1** displays the resulting dendrogram, with each of the 20 clusters shown in a
different color.
```{r fig01, fig.cap = "Figure 1: Hierarchical clustering of break point sequences.", fig.width = 12}
plot(seqclust)
```

**Figure 2**  contains a simplified view of the same dendrogram, cutting off the clutter
below the major splits into branches.
```{r fig02, fig.width=8, fig.cap = "Figure 2: Uncluttered branches."}
plot(seqclust, type = "clipped")
```

**Figure 3** displays an alternate view of the dendrogram, emphasizing the relationships
between clusters. 
```{r fig03, fig.cap = "Figure 3: Unrooted phylogeny tree.", fig.width=9, fig.height=9}
plot(seqclust, type = "unrooted")
```

Both **Figure 2** and, more strongly, **Figure 3** indicate that there are around four
major branches.

We can also show the dendrogram attached to a heatmap containing the distance matrix.

```{r fig04, fig.cap = "Figure 4: Heatmap of the distance matrix.", fig.width=9, fig.height=9}
heat(seqclust)
```

# Aligning Clustered Sequences
Each of the clusters uses a small enough set of symbols that we can safely rewrite them as
though they are sequences of amino acids. We then use the `align` function from the `SVAlignR`
package to perform multiple sequence alignments, using teh "ClustalW" algorithm on each cluster.
Because the components are not amino acids, however, we do not need to use one of the PAM or
BLOSUM matrices. Instead, we define a substiuttion matrix that gives the same match and mismatch
scores to each pair of symbols.

```{r do.align}
MYSUB <- SVAlignR:::makeMYSUB(match = 2, mismatch = -6)
NK <- seqclust@NC
typer <- seqclust@clusters

consensus <- lapply(1:NK, function(K) {
  alignBranch(LR[typer == K], mysub = MYSUB, gapO = 2, gapE = 0.1)
})
names(consensus) <- paste("C", c(rep("0", 9), rep("", 11)), 1:20, sep = "")
class(consensus)
length(consensus)
```

```{r echo=FALSE, eval=FALSE}
all(names(weights) == names(seqclust@clusters))
daft <- data.frame(weights, seqclust@clusters)
daft[weights == 2,]
```

**Figure 5** contains an example of a multiple sequence alignment.
```{r fig05 , fig.cap="Figure 5: Example alignment of cluster 15."}
opar <- par(mai = c(1.02, 0.82, 0.82, 0.82))
wt <- weights[colnames(consensus[[15]]$meet)]
showme(consensus[[15]])
mtext(paste("n =", wt), side = 4, line = 1, at = 1:length(wt), las=2)
par(opar)
rm(opar)
```

## Alignments By Cluster
*Here**Figure 6** shows all 20 alignments. 
```{r fig06, fig.cap="Figure 6: Multiple sequence alignments of clusters.", fig.width=20, fig.height=20}
opar <- par(mfrow = c(4,3))
for (K in 1:NK) {
  showme(consensus[[K]], col = SVAlignR:::myColorSet[K])
}
par(opar)
```

## Consensus Sequences

```{r conseq, fig.cap="Alignment of all consensus sequences.", echo=FALSE, eval = FALSE}
allmodcons <- sapply(consensus, function(X) X$cons)
allmodcons <- gsub("\\?", "q", allmodcons) # make question mark a letter
allmodcons <- gsub("\\:", "z", allmodcons) # make rare colon a letter
while (length(grep("^[qz]", allmodcons) > 0)) {
  allmodcons <- sub("^q-", "", allmodcons)
  allmodcons <- sub("^z-", "", allmodcons)
}
while (length(grep("[qz]$", allmodcons) > 0)) {
  allmodcons <- sub("-q$", "", allmodcons)
  allmodcons <- sub("-z$", "", allmodcons)
}
names(allmodcons) <- paste("C", c(rep(0,9), rep("", 11)), 1:20, sep = "")

newsub <- SVAlignR:::makeMYSUB(match = 2, mismatch = -6)
ab <- alignBranch(allmodcons, mysub = newsub, gapO = 0.1, gapE = 0.2)
showme(ab)

```


```{r fig07, fig.cap = "Figure 7: Unrooted phylogeny tree with consensus sequences.", fig.width=9, fig.height=9}
library(dendextend)
myColorSet <- SVAlignR:::myColorSet
NC <- seqclust@NC
mycut <- mean(rev(seqclust@hc$height)[NC + -1:0])
hcd <- as.dendrogram(seqclust@hc)
dend2 <- cut(hcd, h = mycut)
DU <- dend2$upper
blocks <- sapply(consensus, function(X) {
  X$cons
})
labels(DU) <- paste("   [", 1:NC, "]  ",  blocks, "   ", sep = "")

opar <- par(bg = "gray90")
plot(ape::as.phylo(DU), type = "unrooted", rotate.tree = -20,
     edge.color = "black", edge.width = 2,
     lab4ut = "axial", 
     tip.color = myColorSet[1:NK], cex = 1.2)
par(opar)

```

```{r makepngs, eval = FALSE, echo=FALSE, results = "hide"}
resn <- 300
for (K in 1:NK) {
  temp <- consensus[[K]]
  png(file = file.path(paths$figs, paste("cons-", K, ".png", sep = "")),
      width = 6*resn, height = 4*resn, res = resn, bg = "white")
  showme(temp, col = SVAlignR:::myColorSet[K], cex = 0.5)
  dev.off()
}
```

# Appendix
```{r si}
sessionInfo()
```